---
title: "Practical-Machine-Learning-July-2014-LeifUlstrup.Rmd"
author: "Leif Ulstrup"
date: "July 24, 2014"
output: html_document
---
##Executive Summary

##Context

Source of Experimental Data: http://groupware.les.inf.puc-rio.br/har (see section "Weight Lifting Exercises Dataset"").  See this paper: http://groupware.les.inf.puc-rio.br/work.jsf?p1=11201.  From the paper:
"...Participants were asked to perform one set of 10 repetitions
of the Unilateral Dumbbell Biceps Curl in Five different fashions:
exactly according to the specification (Class A), throwing
the elbows to the front (Class B), lifting the dumbbell
only halfway (Class C), lowering the dumbbell only halfway
(Class D) and throwing the hips to the front (Class E). Class
A corresponds to the specified execution of the exercise,
while the other 4 classes correspond to common mistakes..."

##Model Development

###How the Model was Built

###How Cross-validation was Used

###Expected Sample Error Calculations

###Reasoning Behind Choices Made

###Link to Predictions

###R Code

```{r loadLibraries, echo=FALSE}
require(caret)

install.packages("doMC") # install parallel processing library
require(doMC)
registerDoMC(cores = 4)

require("randomForest") # for model training builds

```

```{r ingestData, echo=FALSE}

#assume that the working directory has been set correctly
#download source files to current working directory
# commented out for this effort since they are already downloaded (remove otherwise to test)
#download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", "pml-training.csv", method="curl")
#download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", "pml-testing.csv", method="curl")
if(!("pml-testing.csv" %in% list.files())) stop("Code Stopped: Missing testing.csv file")
if(!("pml-training.csv" %in% list.files())) stop("Code Stopped: Missing testing.csv file")

#test files are in the directory

trainingFile <- read.csv("pml-training.csv") # source file for training and testing fit
testFile <- read.csv("pml-testing.csv") # ultimate test file for project submission

```

```{r dataCleanup, echo=FALSE}
# remove nearZero Values and NAs

# match the columns in the training source to those used in the test set
keepcols <- c()
for (i in names(testFile)) {
  if(!is.na(testFile[[i]][1])){
    #print(paste(i, "is NOT NA"))
    keepcols <- c(keepcols, i)
    }
  }

trainingClean <- data.frame()
trainingClean <- trainingFile[, (names(trainingFile) %in% keepcols)]
trainingClean$classe <- trainingFile$classe # indicator of how well the bar bell movement is performed on a scale of A to E

testClean <- data.frame()
testClean <- testFile[, (names(trainingFile) %in% keepcols)]
testClean$problem_id <- testFile$problem_id # note that the last column is "problem_id" instead of "classe"

trainingClean <- trainingClean[-c(1:7)] #remove the names, timestamp etc columns
testClean <- testClean[-c(1:7)] #remove the names, timestamp etc columns

#convert to "numeric" class except for the classe factor
for (i in 1:(length(trainingClean)-1)) {trainingClean[[i]] <- as.numeric(trainingClean[[i]])}
for (i in 1:(length(testClean)-1)) {testClean[[i]] <- as.numeric(testClean[[i]])}

```

```{r createTrainingandTest, echo=FALSE}

inTrain <- createDataPartition(y=trainingClean$classe, p=0.7, list=FALSE) # Note use of "clean" version of trainingFile
training <- trainingClean[inTrain,] # Note use of "clean" version of trainingFile
testing <- trainingClean[-inTrain,] # Note use of "clean" version of trainingFile

# !!! huge roadblock solved - another clean up item (per Class Forum post)
#  see this stackoverflow discussion:
# http://stackoverflow.com/questions/13495041/random-forests-in-r-empty-classes-in-y-and-argument-legth-0
training$classe <- factor(training$classe) #key cleanup actvity
testing$classe <- factor(testing$classe) #key cleanup actvity

# we are now set up with a set of data we can train and test our models on...let's try some models

```

```{r analyzeCorrelations, echo=FALSE}

trainingCor <- cor(training[, -length(training)]) # account for the classe factor on last column
highCor <- sum(abs(training[upper.tri(trainingCor)]) > 0.999)

highlyCor <- findCorrelation(training[, -length(training)], cutoff = 0.75)

preProc <- preProcess(training[, -length(training)], method= c("center", "scale"))

trainPredict <- predict(preProc, newdata = training[, -length(training)])

```

```{r modeling, echo=FALSE}
set.seed(12345)

# let's try a model using 

preProc <- preProcess(training[,-53], method="pca") # "-53" removes the $classe factor element
trainPC <- predict(preProc, training[,-53]) # note again extraction of $classe
modelFit <- train(training$classe ~ ., method="rf", data=trainPC) # strange construct, rf for RandomForest
print(modelFit)
testPC <- predict(preProc, testing[,-53])
confusionMatrix(testing$classe, predict(modelFit, testPC))


```
